\propertitle

### 2019-12-20 maxNt_v6.tex
update

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-12-19T20:55:39+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Inferring the total activity of a large neuronal population\\ from a small sample}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Inferences about large neuronal populations}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

This fact affects the interpretation of the results obtained from
maximum-entropy methods. For example, if means and pairwise correlations
seem informationally sufficient for a particular sample from a brain area,
then they are probably not sufficient for the larger population of neurons
constituting that area, and vice versa. If we are interested in the
cooperativity of the neurons in a brain area, it is therefore unreliable to
use a maximum-entropy distribution constructed only for a sample
thereof. % \citep[contrast this with some literature, \eg][]{martignonetal1995,bohteetal2000,schneidmanetal2006,shlensetal2006,barreiroetal2010,ganmoretal2011,granotatedgietal2013}
The model presented here avoids this problem because the maximum-entropy
method is applied directly to the larger population, not to the sample
alone.

@@

Let us finally illustrate the idea and remarks above with our first data set.

We measure the difference $\ydi(M'',M')$ in informational sufficiency
between a set of moments, say $M''\defd\set{1,\dotsc,m''}$, and another, say
$M'\defd \set{1,\dotsc,m'}$, as follows:
\begin{enumerate}[label=(\roman*)]
\item From each maximum-entropy distribution $\px(\yA \| M,N)$ for the
  larger population, built from each set of constraints $M=M',M''$,
  calculate the marginal distribution for the sample by
  \eqn~\eqref{eq:conditional_hypergeometric_t}:
\begin{equation}
  \label{eq:sample_marginal}
  \pxx(\ya \| M, N) = \sum_{\yA}\yG_{\ya \yA}\px(\yA \| M,N),
  \qquad M=M',M''.
\end{equation}
\item Calculate the relative entropies of the measured frequency
  distribution $\yf$ with respect to each sample marginal from the previous
  step, and multiply them by the number of time bins $T$:
  \begin{equation}
    \label{eq:relentropy_T}
    T\,\sh[\yf; \pxx(\ya \| M,N)] \defd
    T  \sum_{\ya} \yff_{\ya} \log\frac{\yff_{\ya}}{\pxx(\ya \| M,N)},
    \qquad M=M',M''.
  \end{equation}
\item Take the difference: 
\begin{equation}
  \label{eq:measure_suffic}
  \begin{split}
  \ydi_{N}(M'',M') &\defd
  T \,\sh[\yf; \pxx(\ya \| M',N)]
  -
  T \,\sh[\yf; \pxx(\ya \| M'',N)]
  \\ &\equiv
T  \sum_{\ya} \yff_{\ya} \log\frac{
  \sum_{\yA}\yG_{\ya\yA}\px(\yA \| M'',N)}{
  \sum_{\yA}\yG_{\ya\yA}\px(\yA \| M',N)}.
\end{split}
\end{equation}
\end{enumerate}
The measure $\ydi_{N}(M'',M')$ so defined is positive if $M''$ is more
probable to be informationally sufficient than $M'$, and negative
otherwise.

We show in appendix~\ref{sec:measure_suff} that $\ydi_{N}(M'',M')$ is equal
to the log-ratio of the probabilities of the data $\yf$ conditional on the
hypotheses $M''$ and $M'$:
\begin{equation}
  \label{eq:bayes_factor}
  \ydi_{N}(M'',M') =
  \log\frac{\pf(\yf \| M'',N)}{\pf(\yf \| M',N)}.
\end{equation}
This is their \emph{relative weight of evidence}
\citep[\chap~6]{good1950}{good1985}[\sect~1.4]{osteyeeetal1974}, the
logarithm of their \emph{relative Bayes factor}
\citep[p.~421]{jeffreys1936}[and references
therein]{kassetal1995}. %[\chaps~V, VI, A]{jeffreys1939_r1983}.
The exponential of $\ydi_{N}(M'',M')$ tells us how much more probable the
data $\yf$ are, conditional on the moments $M''$, than conditional on the
moments $M'$. We can also combine this measure with pre-data probabilities
for the two hypotheses to obtain the ratio of their probabilities
conditional on the data \citep[\cf][]{bretthorst2013}.

@@

Let's show a concrete application. Consider for example three sets of
constraints $M_{2}$, $M_{4}$, $M_{5}$, consisting of the first two, four,
five moments; and a larger population of size $N=10\,000$.
Figure~\ref{fig:plots_2_4mom} shows the resulting distributions for the two
data sets.
\begin{figure}[!p]
\centering
\subcaption[]{\label{fig:plots_2_4mom_a}}%  
\includegraphics[width=\linewidth]{_scripts/newpaper2_stensola_threemom_distributions10000_priorc.pdf}%
\\[2em]%
\subcaption[]{\label{fig:plots_2_4mom_b}}%  
\includegraphics[width=\linewidth]{_scripts/newpaper2_riehle_threemom_distributions10000_prior2.pdf}%
\\[2em]%
\caption{Maximum-entropy distributions for the normalized total activity of
  the larger population, assuming two different sets of constraining moments
  and a larger population size $N=10\,000$.}
\label{fig:plots_2_4mom}
\end{figure}
% maxent_solutions_different_moments_prior2_6moments.nb
There is a clear visual difference between the distributions based on two
(\textcolor{myred}{dashed red}) and on four (\textcolor{myblue}{solid
  blue}) moments; for the first data set, panel
\subcaptionref{fig:plots_2_4mom_a}, they even differ in the number of
modes. On the other hand there is barely any visual difference between the
distributions based on four and on five (\textcolor{myyellow}{dotted
  yellow}) moments. From these visual differences we immediately conclude
that two moments are probably not informationally sufficient, whereas four
and five should both be. We shall see that the
measure~\eqref{eq:measure_suffic} reflects what we immediately see.

Compare these plots with those obtained applying the model at the sample
level (equivalent to $N=n$). With the $M_{2}$ and $M_{4}$ constraints we
obtain the distributions partly shown in
\fig~\ref{fig:comparisons_marginals} as \textcolor{myred}{red empty
  circles} for two moments, and \textcolor{myblue}{blue empty squares} for
four moments. The measured frequencies (black triangles) are also shown.
The two distributions can only be distinguished on a logarithmic scale, and
only in their tails. It would be difficult to visually assess which is more
probable to be informationally sufficient.

@@

It is in fact possible to use the measure~\eqref{eq:measure_suffic} and its
probabilistic interpretation~\eqref{eq:bayes_factor} to compare the
probability of the data $\yf$ conditional on the hypothesis $(M_{4},N)$ of
four-moment sufficiency at the larger-population level, $N=10\,000$, versus
the hypothesis $(M_{4}, n)$ of four-moment sufficiency at the sample level,
$N=n=65$. We obtain
\begin{equation}
  \label{eq:diff_suff_n}
  \ydi[(M_{4}, N), (M_{4}, n)] = 39\;\nat = 17\;\hart.
\end{equation}
% that is, the data are 17 orders of magnitude more probable under the first
% hypothesis than under the second. 
This result is also shown in
\fig~\ref{fig:comparisons_marginals}, where we see that the four-moment
distribution constructed at the larger-population level
(\textcolor{myblue}{blue filled squares}) fits the measured frequency
distribution (black triangles) more closely than the corresponding
sample-level distribution (\textcolor{myblue}{blue empty squares}).

\bigskip
\mynote{Short reasoned bibliography about sufficient statistics and its
  relation to maximum-entropy?} blah blah \footnote{For more technical results and
  connections with the notion of symmetry see \eg\
  \cite{darmois1935,neyman1935,koopman1936,pitman1936,halmosetal1949,bahadur1954,berk1972,lauritzen1974,lauritzen1982_r1988,lauritzen2007,cifarellietal1980,cifarellietal1981,diaconisetal1981,diaconis1992,furmanczyketal1998,fortinietal2000,nogalesetal2000,kallenberg2005,ayetal2015}.}
This mathematical result comes from the Pitman-Koopman
theorem\citep{cifarellietal1982,kullbacketal1951}[the
notion goes back to][]{fisher1922} \citep{jaynes1982b}[\sect~4.5.4]{bernardoetal1994}
\citep{koopman1936,pitman1936,darmois1935}[for later analyses and the
discrete case
see][]{hipp1974,andersen1970,denny1967,denny1972,fraser1963,barankinetal1963,barndorffnielsen1978_r2014}

>>>>

\pdfoutput=1
%% Author: PGL  Porta Mana
%% Created: 2015-05-01T20:53:34+0200
%% Last-Updated: 2019-12-20T11:10:29+0100
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\newif\ifarxiv
\arxivfalse
\ifarxiv\pdfmapfile{+classico.map}\fi
\newif\ifafour
\afourfalse% true = A4, false = A5
\newif\iftypodisclaim % typographical disclaim on the side
\typodisclaimtrue
\newcommand*{\memfontfamily}{zplx}
\newcommand*{\memfontpack}{newpxtext}
\documentclass[\ifafour a4paper,12pt,\else a5paper,10pt,\fi%extrafontsizes,%
onecolumn,oneside,article,%french,italian,german,swedish,latin,
british%
]{memoir}
\newcommand*{\updated}{\today}
\newcommand*{\firstdraft}{4 November 2015}
\newcommand*{\firstpublished}{***}
\newcommand*{\propertitle}{Inferring the total activity of a large neuronal population\\ from a small sample}
\newcommand*{\pdftitle}{\propertitle}
\newcommand*{\headtitle}{Inferences about large neuronal populations}
\newcommand*{\pdfauthor}{P.G.L.  Porta Mana, Y. Roudi, V. Rostami, E. Torre}
\newcommand*{\headauthor}{Porta Mana \etal}
\newcommand*{\reporthead}{}% Report number

@@

This fact affects the interpretation of the results obtained from
maximum-entropy distributions. For example, if means and pairwise correlations
seem informationally sufficient for a particular sample from a brain area,
then they are probably not sufficient for the larger population of neurons
constituting that area, and vice versa. If we are interested in the
cooperativity of the neurons in a brain area, it is therefore unreliable to
use a maximum-entropy distribution constructed only for a sample
thereof. % \citep[contrast this with some literature, \eg][]{martignonetal1995,bohteetal2000,schneidmanetal2006,shlensetal2006,barreiroetal2010,ganmoretal2011,granotatedgietal2013}
The model presented here avoids this problem because the maximum-entropy
method is applied directly to the larger population, not to the sample
alone.

@@

We compare the informational sufficiency between of a set of moments, say
$M''\defd\set{1,\dotsc,m''}$, with respect to another, say
$M'\defd \set{1,\dotsc,m'}$, by introducing the quantity $\ydi(M'',M')$
defined as follows:
\begin{enumerate}[label=(\roman*)]
\item From each maximum-entropy distribution $\px(\yA \| M,N)$ for the
  larger population, built from each set of constraints $M=M',M''$,
  calculate the marginal distribution for the sample by
  \eqn~\eqref{eq:conditional_hypergeometric_t}:
\begin{equation}
  \label{eq:sample_marginal}
  \pxx(\ya \| M, N) = \sum_{\yA}\yG_{\ya \yA}\px(\yA \| M,N),
  \qquad M=M',M''.
\end{equation}
\item Calculate the relative entropies of the measured frequency
  distribution $\yf$ with respect to each sample marginal from the previous
  step, and multiply them by the number of time bins $T$:
  \begin{equation}
    \label{eq:relentropy_T}
    T\,\sh[\yf; \pxx(\ya \| M,N)] \defd
    T  \sum_{\ya} \yff_{\ya} \log\frac{\yff_{\ya}}{\pxx(\ya \| M,N)},
    \qquad M=M',M''.
  \end{equation}
\item Take the difference: 
\begin{equation}
  \label{eq:measure_suffic}
  \begin{split}
  \ydi_{N}(M'',M') &\defd
  T \,\sh[\yf; \pxx(\ya \| M',N)]
  -
  T \,\sh[\yf; \pxx(\ya \| M'',N)]
  \\ &\equiv
T  \sum_{\ya} \yff_{\ya} \log\frac{
  \sum_{\yA}\yG_{\ya\yA}\px(\yA \| M'',N)}{
  \sum_{\yA}\yG_{\ya\yA}\px(\yA \| M',N)}.
\end{split}
\end{equation}
\end{enumerate}
The quantity $\ydi_{N}(M'',M')$ so defined is positive if $M''$ is more
probable to be informationally sufficient than $M'$, and negative
otherwise.

We show in appendix~\ref{sec:measure_suff} that $\ydi_{N}(M'',M')$ is
approximately equal to the log-ratio of the probabilities of the data $\yf$
conditional on the hypotheses $M''$ and $M'$:
\begin{equation}
  \label{eq:bayes_factor}
  \ydi_{N}(M'',M') \approx
  \log\frac{\pf(\yf \| M'',N)}{\pf(\yf \| M',N)}.
\end{equation}
This is their \emph{relative weight of evidence}
\citep[\chap~6]{good1950}{good1985}[\sect~1.4]{osteyeeetal1974}, the
logarithm of their \emph{relative Bayes factor}
\citep[p.~421]{jeffreys1936}[and references
therein]{kassetal1995}. %[\chaps~V, VI, A]{jeffreys1939_r1983}.
The exponential of $\ydi_{N}(M'',M')$ tells us how much more probable the
data $\yf$ are, conditional on the sufficiency of $M''$, than conditional
on the sufficiency of $M'$. We can also combine this measure with pre-data
probabilities for the two hypotheses to obtain the ratio of their
probabilities conditional on the data \citep[\cf][]{bretthorst2013}.

@@

Let us illustrate the ideas, the remarks, and the quantity defined so far
with our first data set. Consider for example three sets of constraints
$M_{2}$, $M_{4}$, $M_{5}$, consisting of the first two, four, five moments;
and a larger population of size $N=10\,000$. Figure~\ref{fig:plots_2_4mom}
shows the resulting distributions for the two data sets.
\begin{figure}[!p]
\centering
\subcaption[]{\label{fig:plots_2_4mom_a}}%  
\includegraphics[width=\linewidth]{_scripts/newpaper2_stensola_threemom_distributions10000_priorc.pdf}%
\\[2em]%
\subcaption[]{\label{fig:plots_2_4mom_b}}%  
\includegraphics[width=\linewidth]{_scripts/newpaper2_riehle_threemom_distributions10000_prior2.pdf}%
\\[2em]%
\caption{Maximum-entropy distributions for the normalized total activity of
  the larger population, assuming two different sets of constraining moments
  and a larger population size $N=10\,000$.}
\label{fig:plots_2_4mom}
\end{figure}
% maxent_solutions_different_moments_prior2_6moments.nb
There is a clear visual difference between the distributions based on two
(\textcolor{myred}{dashed red}) and on four (\textcolor{myblue}{solid
  blue}) moments; for the first data set, panel
\subcaptionref{fig:plots_2_4mom_a}, they even differ in the number of
modes. On the other hand there is barely any visual difference between the
distributions based on four and on five (\textcolor{myyellow}{dotted
  yellow}) moments. From these visual differences we immediately conclude
that two moments are less likely to be informationally sufficient than
four; but four should almost as likely as five. We shall see that the
measure~\eqref{eq:measure_suffic} reflects what we immediately see.

Compare these plots with those obtained applying the model at the sample
level (equivalent to $N=n$). With the $M_{2}$ and $M_{4}$ constraints we
obtain the distributions partly shown in
\fig~\ref{fig:comparisons_marginals} as \textcolor{myred}{red empty
  circles} for two moments, and \textcolor{myblue}{blue empty squares} for
four moments. The measured frequencies (black triangles) are also shown.
The two distributions can only be distinguished on a logarithmic scale, and
only in their tails. It would be difficult to visually assess which is more
likely to be informationally sufficient.

@@

It is in fact possible to use the measure~\eqref{eq:measure_suffic} and its
probabilistic interpretation~\eqref{eq:bayes_factor} to compare the
probability of the data $\yf$ conditional on the hypothesis $(M_{4},N)$ of
four-moment sufficiency at the larger-population level, $N=10\,000$, versus
the hypothesis $(M_{4}, n)$ of four-moment sufficiency at the sample level,
$N=n=65$. We obtain
\begin{equation}
  \label{eq:diff_suff_n}
  \ydi[(M_{4}, N), (M_{4}, n)] = 39\;\nat = 17\;\hart.
\end{equation}
% that is, the data are 17 orders of magnitude more probable under the first
% hypothesis than under the second. 
This result is also shown in
\fig~\ref{fig:comparisons_marginals}. We see that the four-moment
distribution constructed at the larger-population level
(\textcolor{myblue}{blue filled squares}) fits the measured frequency
distribution (black triangles) more closely than the corresponding
sample-level distribution (\textcolor{myblue}{blue empty squares}).

We close this section with some references. The notion of informational or
predictive sufficiency, probably first presented by Kullback \amp\ Leibler
\parencite*{kullbacketal1951}, is discussed in Bernardo \etal\
\parencite*[\sect~4.5]{bernardoetal1994}, who explain its connections with
Fisher's\parencite*{fisher1922} parametric sufficiency. Jaynes
\parencite*[\chap~8 and \sect~14.2] shows the connections with information
theory. The mathematical relation between sufficiency and the exponential
family comes from the Pitman-Koopman
theorem \citep{koopman1936,pitman1936,darmois1935} for continuous variables.

\mynote{Short reasoned bibliography about sufficient statistics and its
  relation to maximum-entropy?} blah blah \footnote{For more technical results and
  connections with the notion of symmetry see \eg\
  \cite{darmois1935,neyman1935,koopman1936,pitman1936,halmosetal1949,bahadur1954,berk1972,lauritzen1974,lauritzen1982_r1988,lauritzen2007,cifarellietal1980,cifarellietal1981,diaconisetal1981,diaconis1992,furmanczyketal1998,fortinietal2000,nogalesetal2000,kallenberg2005,ayetal2015}.}
This mathematical result comes from the Pitman-Koopman
theorem\citep{cifarellietal1982,kullbacketal1951}[the
notion goes back to][]{fisher1922} \citep{jaynes1982b}[\sect~4.5.4]{bernardoetal1994}
\citep{koopman1936,pitman1936,darmois1935}[for later analyses and the
discrete case
see][]{hipp1974,andersen1970,denny1967,denny1972,fraser1963,barankinetal1963,barndorffnielsen1978_r2014}
